Multi-stage programming with functors and monads: eliminating abstraction overhead from generic code

### 2008-06-12 SCP2008/paper/metamonads.tex
Lots of work, especially complete rewriting of Sec on lifted domains

\usepackage{fancyvrb}
\DefineShortVerb{\|}
\DefineVerbatimEnvironment{code}{Verbatim}{xleftmargin=\ourmathindent,commandchars=\\\{\},fontsize=\small}
\DefineVerbatimEnvironment{code2}{Verbatim}{xleftmargin=\ourmathindent,commandchars=\\\{\},fontsize=\scriptsize}
\newcommand{\evalresult}[1]{\ensuremath{\Longrightarrow}\textcolor{red}{#1}}
% \setlength{\parskip}{0pt}
\parskip=0pt
\newlength{\ourmathindent}
\setlength{\ourmathindent}{1em}

@@

\subsection{Aspects}

@@

\subsection{Organizing aspects}

@@

\begin{enumerate}
    \item \textbf{Abstract domain}.  This group includes
      `mathematical' aspects: domain of matrix elements, length
      measure, full division, normalization, zero testing,
      symmetry and other matrix structure.
    \item \textbf{Concrete representation}: choosing data
      structures for domains, containers, permutation
      matrices. This group also includes packing, in-place
      decomposition, and the
      representation of sparse matrices.
    \item \textbf{Interface} of each generated function, including
      the input and output choices, logging and error reporting.
    \item \textbf{Algorithmic Strategy}, such as fraction-free
      updates, pivoting strategies, augmented matrices.
    \item \textbf{Tracking}, of determinant, rank, or pivot, etc.
    \item \textbf{Staging}: whether the result is the program or a
      generator that will produce the program.
\end{enumerate}

@@

\begin{code}
let one = .<1>. and plus x y = .<.~x + .~y>.
let simplest_code = let gen x y = plus x (plus y one) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
\evalresult{.<fun x_1 -> fun y_2 -> (x_1 + (y_2 + 1))>.}
\end{code}

@@

Let us consider a more complex expression:
\begin{code}
let param_code1 plus one =
  let gen x y = plus (plus y one) (plus x (plus y one)) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
\end{code}
with two occurrences of |plus y one|,
which may be a rather complex computation which we would rather not do
twice. We may be tempted to rely on the compiler's
common-subexpression elimination optimization. When the generated code is
very complex, however, the compiler may overlook common subexpressions.  Or the
subexpressions may occur in an imperative context where the compiler
might not be able to determine whether lifting them is sound. So, being
conservative, the optimizer will leave the duplicates as they are. 
We may attempt to eliminate subexpressions as follows: 
\begin{code}
let param_code1' plus one =
  let gen x y = let ce = (plus y one) in  plus ce (plus x ce) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
param_code1' plus one
\evalresult{.<fun x_1 -> fun y_2 -> ((y_2 + 1) + (x_1 + (y_2 + 1)))>.}
\end{code}
However,
the result of |param_code1' plus one| still exhibits duplicate
sub-expressions.  Our |let|-insertion optimization only saved the
computation at the generating stage.  We need a combinator that
inserts the |let| expression in the generat\emph{ed} code. We need a
combinator |letgen| to be used as
\begin{code}
let ce = letgen (plus y one) in plus ce (plus x ce)
\end{code}
yielding code like 
\begin{code}
.<let t = y + 1 in t + (x + t)>.
\end{code}
But that seems impossible because |letgen exp| has to generate
the expression |.<let t = exp in body>.| but |letgen| does not yet
have the |body|. The body needs a temporary identifier |.<t>.|
that is supposed to be the result of |letgen| itself.  Certainly
|letgen| cannot generate only part of a let-expression, without the
|body|, as all generated expressions in MetaOCaml are well-formed and
complete.

@@

\begin{code}
let letgen exp k = .<let t = .~exp in .~(k .<t>.)>.
let param_code2 plus one =
  let gen x y k = letgen (plus y one)
                         (fun ce -> k (plus ce (plus x ce)))
  and k0 x = x
  in .<fun x y -> .~(gen .<x>. .<y>. k0)>.
param_code2 plus one
\evalresult{.<fun x_1 -> fun y_2 -> let t_3 = (y_2 + 1) in (t_3 + (x_1 + t_3))>.}
\end{code}

@@

\subsection{Generating control statements}
We can write operations that generate code other than let-statements,
e.g., conditionals: see |ifL| in Figure~\ref{ourmonad}. The function |ifL|, 
albeit straightforward, is not as general as we wish: its arguments are
already generated pieces of code rather than monadic values. We
can ``lift it'':
\begin{code}
let ifM' test th el = perform
  testc <-- test; thc <-- th; elc <-- el;
  ifL testc thc elc
\end{code}
However we also need another |ifM| function, with the same
interface (see Figure~\ref{ourmonad}).
The difference between them is apparent from the following example:
\begin{code}
let gen a i = ifM' (ret .<(.~i) >= 0>.) 
                   (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
 in .<fun a i -> .~(runM (gen .<a>. .<i>.) ())>.
\evalresult{.<fun a_1 i_2 ->}
\textcolor{red}{      let t_3 = (Some a_1.(i_2)) in if (i_2 >= 0) then t_3 else None>.}
let gen a i = ifM (ret .<(.~i) >= 0>.) 
                  (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
 in .<fun a i -> .~(runM (gen .<a>. .<i>.) ())>.
\evalresult{.<fun a_1 i_2 ->}
\textcolor{red}{      if (i_2 >= 0) then let t_3 = (Some a_1.(i_2)) in t_3 else None>.}
\end{code}
%
If we use |ifM'| to generate guarded array access code, the let-insertion
happened \emph{before} the if-expression, that is, before the test that
the index |i| is positive. If |i| turned out
negative, |a.(i)| would generate an out-of-bound array access
error. On the other hand, the code with |ifM| accesses the array only
when we have verified that the index is non-negative. This example
makes it clear that code generation (such as the one in |retN|) is 
truly an effect, and we have to be clear about the sequencing of
effects when generating control constructions such as conditionals.
The form |ifM| handles such effects correctly.

@@

Various aspects of our generator will need to keep a state during
code generation (see \S\ref{sec:det} for an example).

@@

The MLton team suggests a better approach: property 
lists~\cite{mlton-proplist}.  A property list also represents an
extensible object but by its dual, namely a list of polymorphic variants.  The
initial object is just the empty list.  Unfortunately, we cannot apply
MLton's approach litterally to our case.  MetaOCaml's environment classifiers
require type universes parameterized by classifiers. The MLton
approach uses generativity of exceptions or reference cells to
generate property names, i.e., fields of an extensible record. Using this
technique would make our aspects stateful modules, something we do not
want, as we wish our generator to be purely functional [even though it is
generating imperative code].
Fortunately, OCaml offers a way to build open unions with
polymorphic variants. Each variant is identified by a manifest tag,
e.g., |`Tdet|, and may include a value. The tags are not generative and
provide a form of manifest naming, similar to symbols of
Lisp and Scheme.

@@

For detailed reference, we present one particular Gaussian Elimination
algorithm, for an in-place LU-decomposition of an integer matrix with
full pivoting, returning the U-factor, determinant and rank. The
$n\times m$-matrix is represented as a flat vector, with 0-indexed
elements laid out in row-major format (C-style). The code below is
hand-written, implementing the typical pseudo-code given in textbooks on
Numerical Analysis \oleg{I used wikipedia; need a better references}.

In OCaml, the matrix is represented by the value of the type
|'a container2dfromvector|:
\begin{code}
type 'a container2dfromvector = {arr:('a array); n:int; m:int}

@@

let swap_cols a (n,m) c j =
  let end_vector = n * m in
  let rec loop col_c col_j =
    if col_j < end_vector then
      begin
        swap a col_c col_j;
        loop (col_c + m) (col_j + m)
      end
  in loop c j
\end{code}
We refer to the $a_{ij}$-the element of the matrix as |a.(i*m+j)|,
where |a| is the array component of the record |container2dfromvector|.
The figure defines three auxiliary functions, typically used in the
textbook pseudo-code: |swap a i j| swaps two elements; 
|swap_cols a (n,m) c j| swaps the column |c| with the column |j| in
the |(n,m)|-matrix |a|. The function 
|swap_rows a (n,m) (r,c) i| swaps the row |r| with the row |i| in 
the non-yet examined portion of the matrix, 
rectangular block |(r,c)-(n,m)|. We do not touch the 
elements to the left of the column |c| because they are all zeros. 
Since we know the layout of the matrix, we avoid 2D index
computations.

@@

let ge = fun a_orig ->
    let r = ref 0 in                    (* current row index, 0-based *)
    let c = ref 0 in                    (* current column index, 0-based *)
    let a = Array.copy (a_orig.arr) in  (* to avoid clobbering A, save it *)
    let m = a_orig.m in                 (* the number of columns *)
    let n = a_orig.n in                 (* the number of rows *)
    let det_sign = ref 1 in             (* Accumulate sign and magnitude *)
    let det_magn = ref 1 in             (*   of the determinant *)
    while !c < m && !r < n do
        (* Look for a pivot *)
        let pivot = find_pivot a (n,m) !r !c in
        let piv_val = (match  pivot with
        | Some ((piv_r, piv_c),piv_val) ->
            if piv_c <> !c then
               begin
                 swap_cols a (n,m) !c piv_c;
                 det_sign := - !det_sign (* flip the sign of the det *)
               end;
            if piv_r <> !r then
               begin
                 swap_rows a (n,m) (!r,!c) piv_r;
                 det_sign := - !det_sign (* flip the sign of the det *)
               end;
            Some piv_val
        | None -> None) in
        (* now do the row-reduction over the (r,c)-(n,m) block *)
        (match piv_val with
        | Some a_rc -> begin
            for ii = !r+1 to n-1 do
                let cur = a.(ii*m + !c) in
                if not (cur == 0)  then
                  begin
                    for j = !c+1 to m-1 do
                      (* fraction-free elimination *)
                      a.(ii*m+j) <- (a.(ii*m+j) * a_rc - a.(!r*m+j) * cur) 
                                    / ! det_magn
                    done;
                    a.(ii*m+ !c) <- 0
                  end;
            done;
            det_magn := a_rc;
            r := !r + 1                 (* advance the rank only if pivot > 0*)
          end
        | None -> det_sign := 0);
        c := !c + 1
    done;
    (* Final result *)
    ({arr=a; n=n; m=m},                 (* The matrix now has the U factor*)
    (if      !det_sign = 0 then  0      (* Compute the signed det *)
     else if !det_sign = 1 then  !det_magn 
     else                        (- !det_magn)),
    !r)                                 (* Rank *)
\end{code}
\caption{Fraction-free in-place Gaussian-elimination over an integer
  matrix}\label{fig:GE-algorithm}
\end{figure}
As typically presented in the textbooks, the main algorithm depends on
the separately-defined function |pivot| to find a pivot. We use
full-pivoting, i.e., searching the non-yet examined portion of the matrix, 
the rectangular block |(r,c)-(n,m)|, for the element with the 
maximum absolute value. The function returns the value of the
found pivot and its location; the return value is an option type with
|None| indicating that all examined elements are zeros. In the main
algorithm, after we found the pivot, we swap the current column 
with the pivot column, and swap the current row with the pivot row. 
After the swap, the $a_{rc}$ element of the matrix is the pivot.
Swapping two rows or two columns changes the sign of the determinant.
After the swapping, the algorithm does so-called row-reduction
over the |(r,c)-(n,m)| block of the matrix containing not-yet
processed elements. We implement the \emph{fraction-free} algorithm:
the division operation in line XXX assuredly divides two integers
without remainder. The fraction-free algorithm requires accumulation
of the determinant \oleg{Citation needed!}.

@@

\jacques{next paragraph will need rewritten when this section has
finally taken shape}
In this section, we will go through various aspects of one algorithm
and explain how they are implemented.  In particular, and generally in
order of increasing complexity, we will deal with abstracting the
domain, the matrix representation, the code itself, determinant, etc.
Then we put it all together (or in a new section?).

@@

We start by abstracting out domains. \oleg{Add: clearly the basic
  structure of the algorithm, Fig.~\ref{fig:GE-algorithm} remains the
  same for integer, float, etc. matrices.} We have already seen the
simplest case of domain abstraction in |param_code1|, which took
code-generators such as |plus| and |one| as arguments.  We need far
more than two parameters: our domains should include $0$, $1$, $+$,
$*$, (unary and binary) $-$, at least \emph{exact} division,
normalization, and potentially a relative size measure. We could group
these parameters in tuples or records.  Instead, we use OCaml
\emph{structures} (i.e., modules) so we can take advantage of
extensibility, type abstraction and constraints, and especially
parameterized structures (\emph{functors}).  We define the type, the
signature |DOMAIN|, which different domains must satisfy:

@@

module IntegerDomain : DOMAIN = struct
  type v = int
  let kind = Domain_is_Ring
  let zero = 0 and one = 1
  let plus x y = x + y  and  minus x y = x - y
  let times x y = x * y and  div x y = x / y
  let uminus x = -x
  let normalizer = None
  let better_than = Some (fun x y -> abs x > abs y)
end
\end{code}

One particular domain instance is |IntegerDomain|. The type annotation
|DOMAIN| in the definition of |IntegerDomain| makes the 
compiler verify that the defined structure is indeed of a type |DOMAIN|. 
\jacques{In fact, it seems that this annotation cannot be put there
because that will make type v abstract, which will result in compilation
errors for the "lifted" domain which needs to have this type be
available.}
\oleg{well, we can say IntegerDomain : DOMAIN with type v = int}
The annotation may be omitted (see |ZpMake| below), in which
case the compiler will verify the type when we try to use that structure as a
|DOMAIN|. In any case, the errors such as missing ``methods'' or
methods with incorrect types will be caught statically, 
\emph{before} any code generation takes place. The variant
|Domain_is_Ring| of |IntegerDomain.domain_kind| encodes a semantic constraint:
that full division
is not available. While the |DOMAIN| type may have looked daunting to
some, the implementation is quite straightforward.  Other domains such
as |float| and arbitrary precision exact rational numbers |Num.num|
are equally simple.

A more complex domain is |Zp|, the field of integers in prime
characteristic:
\begin{code}
module ZpMake(P:sig val p:int end) = struct
    type v = int
    let kind = Domain_is_Field
    let zero = 0 and one = 1
    let plus x y = (x + y) mod P.p
    let times x y = (x * y) mod P.p
    \dots
    let normalizer = None and better_than = None
    let () = assert (is_prime P.p)
end
\end{code}
This domain is parametrized by an integer |p|. 
To be more precise, the structure |ZpMake| is
parameterized over another structure of the type described by
the signature |P|, which has one field, the |int| value |p|.
Such a parameterized structure (or, a function from structures to
structures) is a \emph{functor}. The result of |ZpMake| is a domain which is a
field with no defined order. Hence
|normalizer| and |better_than| are set to |None|. |Zp| forms
a field only when |p| is prime\footnote{or a prime power, a case we do
not treat here.}.  Since we intend to make a field of prime characteristic, 
we must check this, as is done
in the last line of the above code. That line differs from the other 
bindings in |ZpMake| in that it neither defines a function, such as
|plus|, nor binds a value, such as |zero|. This non-value
expression |assert (is_prime P.p)|, which we will call an initializing
expression, will be evaluated
when the corresponding module is instantiated.
\begin{code}
module Z19 = ZpMake(struct let p = 19 end)
\end{code}
If we
replace |p = 19| with |p = 9| above, we receive a run-time
error. However, it is raised as we instantiate and combine modules that
will \emph{eventually} make the generator.  Although the error is reported at
``run-time'' rather than during compilation as one might have hoped, the error
is raised when \emph{generating the generator} -- well before the
generation of the target code could begin.  In our code we make
extensive use of these ``preflight checks'', which are performed as part of
module initialization. These checks seem to offer a good compromise:
they are dynamic and so do not require a complicated type system; on
the other hand, the checks are run quite early, when building
code generators, and so ensure that no code violating the
corresponding \emph{semantic} constraints will be generated. Although
some may frown on the use of module initializing expressions 
as this requires careful attention to sharing and multiple instantiations
of a module, these concerns do not apply in our case: our preflight
checks are all idempotent and maintain no state.

\subsection{Abstract code and lifted domains}\label{sec:lifteddomains}
\jacques{I don't think I really want to delete the first half of this
subsection.  I guess we can move that into section 3?  We need to be
careful that section 3 does not become 1/2 the paper!  We might even
consider splitting section 3 into 2 sections?}
We will be using MetaOCaml operations to build code expressions, values
of type |('a,'v) code|. However, we wish to make our generators
abstract over particular code representations. Therefore, when writing
the LU generator (\S\ref{sec:det} and below), instead of
MetaOCaml brackets and escapes we will be using the code combinators
described below. These combinators are functions that produce and
combine values of the type of \emph{abstract code}.

This encapsulation of staging annotations into code combinators lets
us easily change the output form of our generated LU programs.  For example,
we may use single-stage thunks rather than MetaOCaml staged |('a,'v) code|
values, for benchmarking and regression tests purposes. We could also use
code combinators producing C or Fortran code.

Our code combinator library includes operations to create, dereference
and mutate reference cells, compare values, and manipulate values used
for indexing arrays. The following is the implementation of sample
combinators, in the case of the abstract code being MetaOCaml code. 
We basically need to ``lift'' most the underlying programming language
into combinators.
\begin{code}
let lift x = .< x >.

let cunit = .< () >.  and unitL = fun s k -> k s .< () >.

>>>>

\usepackage[compact]{fancyvrb1}
\DefineShortVerb{\|}
\DefineVerbatimEnvironment{code}{Verbatim}{xleftmargin=\ourmathindent,fontsize=\small}
\DefineVerbatimEnvironment{code2}{Verbatim}{xleftmargin=\ourmathindent,fontsize=\scriptsize}
\newcommand{\evalresult}[1]{\ensuremath{\Longrightarrow}\textcolor{red}{#1}}
% \setlength{\parskip}{0pt}
\parskip=0pt
\newlength{\ourmathindent}
\setlength{\ourmathindent}{1em}

@@

\subsection{Aspects}\label{sec:aspects}

@@

\subsection{Organizing aspects}\label{sec:aspect-groups}

@@

\begin{enumerate}
    \item \textbf{Abstract domain}.  This group includes
      `mathematical' aspects: domain of matrix elements, length
      measure, full division, normalization, zero testing,
      symmetry and other matrix structure.
    \item \textbf{Concrete representation}: choosing data
      structures for domains, containers, permutation
      matrices. This group also includes packing, in-place
      decomposition, and the
      representation of sparse matrices.
    \item \textbf{Interface} of each generated function, including
      the input and output choices, logging and error reporting.
    \item \textbf{Algorithmic Strategy}, such as fraction-free
      updates, pivoting strategies, augmented matrices.
    \item \textbf{Tracking}, of determinant, rank, or pivot, etc.
    \item \textbf{Interpretation}: whether the result is the program or a
      generator that will produce the program.
\end{enumerate}

@@

\begin{code}[commandchars=\\\{\}]
let one = .<1>. and plus x y = .<.~x + .~y>.
let simplest_code = let gen x y = plus x (plus y one) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
\evalresult{.<fun x_1 -> fun y_2 -> (x_1 + (y_2 + 1))>.}
\end{code}

@@

Let us consider a more complex expression:
\begin{code}
let param_code1 plus one =
  let gen x y = plus (plus y one) (plus x (plus y one)) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
\end{code}
with two occurrences of |plus y one|,
which may be a rather complex computation which we would rather not do
twice. We may be tempted to rely on the compiler's
common-subexpression elimination optimization. When the generated code is
very complex, however, the compiler may overlook common subexpressions.  Or the
subexpressions may occur in an imperative context where the compiler
might not be able to determine whether lifting them is sound. So, being
conservative, the optimizer will leave the duplicates as they are. 
We may attempt to eliminate subexpressions as follows: 
\begin{code}[commandchars=\\\{\}]
let param_code1' plus one =
  let gen x y = let ce = (plus y one) in  plus ce (plus x ce) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
param_code1' plus one
\evalresult{.<fun x_1 -> fun y_2 -> ((y_2 + 1) + (x_1 + (y_2 + 1)))>.}
\end{code}
However,
the result of |param_code1' plus one| still exhibits duplicate
sub-expressions.  Our |let|-insertion optimization only saved the
computation at the generating stage.  We need a combinator that
inserts the |let| expression in the generat\emph{ed} code. We need a
combinator |letgen| to be used as
\begin{code}
let ce = letgen (plus y one) in plus ce (plus x ce)
\end{code}
yielding code like 
\begin{code}
.<let t = y + 1 in t + (x + t)>.
\end{code}
But that seems impossible because |letgen exp| has to generate
the expression |.<let t = exp in body>.| but |letgen| does not yet
have the |body|. The body needs a temporary identifier |.<t>.|
that is supposed to be the result of |letgen| itself.  Certainly
|letgen| cannot generate only part of a let-expression, without the
|body|, as all generated expressions in MetaOCaml are well-formed and
complete.

@@

\begin{code}[commandchars=\\\{\}]
let letgen exp k = .<let t = .~exp in .~(k .<t>.)>.
let param_code2 plus one =
  let gen x y k = letgen (plus y one)
                         (fun ce -> k (plus ce (plus x ce)))
  and k0 x = x
  in .<fun x y -> .~(gen .<x>. .<y>. k0)>.
param_code2 plus one
\evalresult{.<fun x_1 -> fun y_2 -> let t_3 = (y_2 + 1) in (t_3 + (x_1 + t_3))>.}
\end{code}

@@

\subsection{Generating control statements}
We can write operations that generate code other than let-statements,
e.g., conditionals: see |ifL| in Figure~\ref{ourmonad}. The function |ifL|, 
albeit straightforward, is not as general as we wish: its arguments are
already generated pieces of code rather than monadic values. We
can ``lift it'':
\begin{code}
let ifM' test th el = perform
  testc <-- test; thc <-- th; elc <-- el;
  ifL testc thc elc
\end{code}
However we also need another |ifM| function, with the same
interface (see Figure~\ref{ourmonad}).
The difference between them is apparent from the following example:
\begin{code}[commandchars=\\\{\}]
let gen a i = ifM' (ret .<(.~i) >= 0>.) 
                   (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
 in .<fun a i -> .~(runM (gen .<a>. .<i>.) ())>.
\evalresult{.<fun a_1 i_2 ->}
\textcolor{red}{      let t_3 = (Some a_1.(i_2)) in if (i_2 >= 0) then t_3 else None>.}
let gen a i = ifM (ret .<(.~i) >= 0>.) 
                  (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
 in .<fun a i -> .~(runM (gen .<a>. .<i>.) ())>.
\evalresult{.<fun a_1 i_2 ->}
\textcolor{red}{      if (i_2 >= 0) then let t_3 = (Some a_1.(i_2)) in t_3 else None>.}
\end{code}
%
If we use |ifM'| to generate guarded array access code, the let-insertion
happened \emph{before} the if-expression, that is, before the test that
the index |i| is positive. If |i| turned out
negative, |a.(i)| would generate an out-of-bound array access
error. On the other hand, the code with |ifM| accesses the array only
when we have verified that the index is non-negative. This example
makes it clear that code generation (such as the one in |retN|) is 
truly an effect, and we have to be clear about the sequencing of
effects when generating control constructions such as conditionals.
The form |ifM| handles such effects correctly.

@@

Various aspects of our generator need to keep a state during
code generation (e.g., the name of the variable for the sign of
the determinant, \S\ref{sec:det}).

@@

The MLton team suggests a better approach: property 
lists~\cite{mlton-proplist}.  A property list also represents an
extensible object but by its dual, namely a list of fields.  The
initial object is just the empty list.  Unfortunately, we cannot apply
MLton's approach literally to our case. The MLton
approach uses generativity of exceptions or reference cells to
generate property names, i.e., fields of an extensible record. This
technique would make our aspects stateful modules, which we described
earlier as undesirable. MLton's approach to generating field
names also makes it difficult to store in the fields MetaOCaml's code
values, whose type contains a generic type variable, the environment
classifier. 
Fortunately, OCaml lets us build open unions with
polymorphic variants. Each variant is identified by a manifest tag,
e.g., |`Tdet|, and may include a value. The tags are not generative and
provide a form of manifest naming, similar to symbols of
Lisp and Scheme.

@@

For detailed reference, we present one particular Gaussian Elimination
algorithm, for an in-place LU-decomposition of an integer matrix with
full pivoting, returning the U-factor, determinant and rank. The
$n\times m$-matrix is represented as a flat vector, with 0-indexed
elements laid out in row-major format (C-style). The code below is
a hand-written implementation of the typical pseudo-code in
Numerical Analysis textbooks \oleg{I used wikipedia; need a better references}.

In OCaml, the matrix is represented by the value of the following type:
\begin{code}
type 'a container2dfromvector = {arr:('a array); n:int; m:int}

@@

let swap_cols a (n,m) c j =
  let end_vector = n * m in
  let rec loop col_c col_j =
    if col_j < end_vector then
      begin
        swap a col_c col_j;
        loop (col_c + m) (col_j + m)
      end
  in loop c j
\end{code}
We refer to the $a_{ij}$-the element of the matrix as |a.(i*m+j)|,
where |a| is the array component of the record |container2dfromvector|.
The above code defines three auxiliary functions, typically used in the
textbook pseudo-code: |swap a i j| swaps two elements; 
|swap_cols a (n,m) c j| swaps the column |c| with the column |j| in
the |(n,m)|-matrix |a|. The function 
|swap_rows| |a| |(n,m)| |(r,c) i| swaps the row |r| with the row |i| in 
the non-yet examined portion of the matrix, 
rectangular block |(r,c)-(n,m)|. We do not touch the 
elements to the left of the column |c| because they are all zeros. 
Since we know the layout of the matrix, we avoid 2D index
computations.

@@

let ge = fun a_orig ->
    let r = ref 0 in                    (* current row index, 0-based *)
    let c = ref 0 in                    (* current column index, 0-based *)
    let a = Array.copy (a_orig.arr) in  (* to avoid clobbering A, save it *)
    let m = a_orig.m in                 (* the number of columns *)
    let n = a_orig.n in                 (* the number of rows *)
    let det_sign = ref 1 in             (* Accumulate sign and magnitude *)
    let det_magn = ref 1 in             (*   of the determinant *)
    while !c < m && !r < n do
        (* Look for a pivot *)
        let pivot = find_pivot a (n,m) !r !c in
        let piv_val = (match  pivot with
        | Some ((piv_r, piv_c),piv_val) ->
            if piv_c <> !c then
               begin
                 swap_cols a (n,m) !c piv_c;
                 det_sign := - !det_sign (* flip the sign of the det *)
               end;
            if piv_r <> !r then
               begin
                 swap_rows a (n,m) (!r,!c) piv_r;
                 det_sign := - !det_sign (* flip the sign of the det *)
               end;
            Some piv_val
        | None -> None) in
        (* now do the row-reduction over the (r,c)-(n,m) block *)
        (match piv_val with
        | Some a_rc -> begin
            for ii = !r+1 to n-1 do
                let cur = a.(ii*m + !c) in
                if not (cur == 0)  then
                  begin
                    for j = !c+1 to m-1 do
                      (* fraction-free elimination *)
                      a.(ii*m+j) <- (a.(ii*m+j) * a_rc - a.(!r*m+j) * cur) 
                                    / ! det_magn
                    done;
                    a.(ii*m+ !c) <- 0
                  end;
            done;
            det_magn := a_rc;
            r := !r + 1                 (* advance the rank only if pivot > 0*)
          end
        | None -> det_sign := 0);
        c := !c + 1
    done;
    (* Final result *)
    ({arr=a; n=n; m=m},                 (* The matrix now has the U factor*)
    (if      !det_sign = 0 then  0      (* Compute the signed det *)
     else if !det_sign = 1 then  !det_magn 
     else                        (- !det_magn)),
    !r)                                 (* Rank *)
\end{code}
\caption{Fraction-free in-place Gaussian-elimination over an integer
  matrix}\label{fig:GE-algorithm}
\end{figure}
As typically presented in the textbooks, the main algorithm depends on
the separately-defined function |pivot| to find a pivot. We use
full-pivoting, i.e., searching the non-yet examined portion of the matrix, 
the rectangular block |(r,c)-(n,m)|, for the element with the 
maximum absolute value. The function returns the value of the
found pivot and its location; the return value is an option type with
|None| indicating that all examined elements are zeros. In the main
algorithm, after we found the pivot, we swap the current column 
with the pivot column, and swap the current row with the pivot row. 
After the swap, the $a_{rc}$ element of the matrix is the pivot.
Swapping two rows or two columns changes the sign of the determinant.
After the swapping, the algorithm does so-called row-reduction
over the |(r,c)-(n,m)| block of the matrix containing not-yet
processed elements. We implement the \emph{fraction-free}
algorithm \oleg{Citation needed!}:
the division operation in line XXX assuredly divides two integers
with no remainder, which requires the accumulation
of the determinant .

@@

In this section, we use the the sample code in
Fig.~\ref{fig:GE-algorithm} to identify (some of the) aspects
discussed in \S\ref{sec:aspects}. We abstract these aspects and
describe their implementation as modules of the code generator.  We
then present the generic Gaussian Elimination algorithm -- what is
left after all the aspects are abstracted away. Finally, we show an
instantiation of the generic generator, whose execution yields code
for the algorithm of Fig.~\ref{fig:GE-algorithm}, this time
automatically generated rather than manually written.

We describe here the following aspects, in the general
order of increasing complexity: domain (of the group `Abstract
domain', see \S\ref{sec:aspect-groups}), code generation combinators
(of the group `Interpretation'), matrix representation (of the group
`Concrete representation'), determinant (the group `Tracking'), and
output (the group `Interface').

@@

Clearly the basic structure of the algorithm,
Fig.~\ref{fig:GE-algorithm}, remains the same for integer, float,
etc. matrices and so can be abstracted over the numeric domain.  We
have already seen the simplest case of domain abstraction in
|param_code1|, which took code-generators such as |plus| and |one| as
arguments.  We need far more than two parameters: our domains should
include $0$, $1$, $+$, $*$, (unary and binary) $-$, at least
\emph{exact} division, normalization, and potentially a relative size
measure. We could group these parameters in tuples or records.
Instead, we use OCaml \emph{structures} (i.e., modules) so we can take
advantage of extensibility, type abstraction and constraints, and
especially parameterized structures (\emph{functors}).  We define the
type, the signature |DOMAIN|, which different domains must satisfy:

@@

module IntegerDomain : DOMAIN with type v = int  = struct
  type v = int
  let kind = Domain_is_Ring
  let zero = 0 and one = 1
  let plus x y = x + y  and  minus x y = x - y
  let times x y = x * y and  div x y = x / y
  let uminus x = -x
  let normalizer = None
  let better_than = Some (fun x y -> abs x > abs y)
end
\end{code}

One particular domain instance is |IntegerDomain|. The type annotation
|DOMAIN| in the definition of |IntegerDomain| makes the 
compiler verify that the defined structure is indeed of a type |DOMAIN|. 
The annotation may be omitted (see |ZpMake| below), in which
case the compiler will verify the type when we try to use that structure as a
|DOMAIN|. In any case, the errors such as missing ``methods'' or
methods with incorrect types will be caught statically, 
\emph{before} any code generation takes place. The variant
|Domain_is_Ring| of |IntegerDomain.domain_kind| encodes a semantic constraint:
that full division
is not available. While the |DOMAIN| type may have looked daunting to
some, the implementation is quite straightforward.  Other domains such
as |float| and arbitrary precision exact rational numbers |Num.num|
are equally simple.

A more complex domain is |Zp|, the field of integers in prime
characteristic:
\begin{code}[commandchars=\\\{\}]
module ZpMake(P:sig val p:int end) = struct
    type v = int
    let kind = Domain_is_Field
    let zero = 0 and one = 1
    let plus x y = (x + y) mod P.p
    let times x y = (x * y) mod P.p
    \dots
    let normalizer = None and better_than = None
    let () = assert (is_prime P.p)
end
\end{code}
This domain is parametrized by an integer |p|. 
To be more precise, the structure |ZpMake| is
parameterized over another structure of the type described by
the signature |P|, which has one field, the |int| value |p|.
Such a parameterized structure (or, a function from structures to
structures) is a \emph{functor}. The result of |ZpMake| is a domain which is a
field with no defined order. Hence
|normalizer| and |better_than| are set to |None|. |Zp| forms
a field only when |p| is prime\footnote{or a prime power, a case we do
not treat here.}.  Since we intend to make a field of prime characteristic, 
we must check this, as is done
in the last line of the above code. That line differs from the other 
bindings in |ZpMake| in that it neither defines a function, such as
|plus|, nor binds a value, such as |zero|. This non-value
expression |assert (is_prime P.p)|, which we will call an initializing
expression, will be evaluated
when the corresponding module is instantiated.
\begin{code}
module Z19 = ZpMake(struct let p = 19 end)
\end{code}
If we
replace |p = 19| with |p = 9| above, we receive a run-time
error. However, it is raised as we instantiate and combine modules that
will \emph{eventually} make the generator.  Although the error is reported at
``run-time'' rather than during compilation as one might have hoped, the error
is raised when \emph{generating the generator} -- well before the
generation of the target code could begin.  In our code we make
extensive use of these ``preflight checks'', which are performed as part of
module initialization. These checks seem to offer a good compromise:
they are dynamic and so do not require a complicated type system; on
the other hand, the checks are run quite early, when building
code generators, and so ensure that no code violating the
corresponding \emph{semantic} constraints will be generated. Although
some may frown on the use of module initializing expressions 
as this requires careful attention to sharing and multiple instantiations
of a module, these concerns do not apply in our case: our preflight
checks are all idempotent and maintain no state.

\subsection{Abstract code and lifted domains}\label{sec:lifteddomains}

In our sample GE code, Fig.~\ref{fig:GE-algorithm}, the operation |not|
|(curr == 0)|, line 46, compares two integers (or, generally, two
domain elements); the operation |det_magn := a_rc|, line 56, assigns
the domain element to the corresponding reference cell; the operation
|r := !r + 1| on the next line increments the rank, the cardinal
number. One can easily imagine a different interpretation of the same
program, where |not (curr == 0)| \emph{generates} code to compare two
domain elements, |det_magn := a_rc| and |r := !r + 1| generate code
for the assignment and the in-place increment. The structure of the GE
algorithm is clearly invariant upon this change in
interpretation. This lets us abstract the algorithm over the
interpretation of basic operations, so that the same GE code, given
different concrete interpretations, can LU-factorize a given matrix,
can generate LU-factorization programs in OCaml, C or Fortran, or can
pretty-print the factorization procedure.

The interpretation aspect is too implemented as an OCaml module.  The
aspect is quite large -- there are many basic operations to abstract
over -- and so the module is structured into several
sub-modules. First we introduce an abstract type |('a,'b) rep| which
describes what sort of objects the interpretation may produce (e.g.,
strings with C code, AST, etc). The type has two parameters: the
second specifies the type of the object, and the first is the
`placeholder' for all other information that may need to be tracked
about the object in a particular interpretation. The interpretation as
MetaOCaml code values uses the first parameter of |rep| to track the
environment classifier.
\oleg{I'm describing in this section coderep.mli and one of its
  instances, code.ml. The only change is that I use ('a,'b) rep rather
  than ('a,'b) abstract, for uniformity and clarity. Why did we even
  need to introduce this additional distinction, `rep' vs
  `abstract'. Also, the name `abstract' is a bit too
  general. Anyway, the explanation does not have to follow the code to
  the letter, as Reviewer 2 has suggested.}
